---
title: "R Syntax Examples"
author: "Nathan Young"
output: pdf_document
---

# R Nuts and Bolts
## Entering Input
We type expressions into the R prompt. 
The <- symbol is the assignment operator
```{r}
x <- 5 # Assignment of value to variable
print(x) # Explicit printing
x # auto-printing
y <- 10:20
```
When [1] is printed it indicates that x is a vector and that 5 is the first element
The : operator is used to create integer sequences

## R Objects
R has five basic classes of objects
* Character
* Numeric (real numbers)
* Integer
* Complex
* Logical (True/False)

The vector is the most basic object type, initialized with vector() function.  
A vector can only contain objects of the same class
A *list* is represented as a vector but can contain objects of different classes. 

## Numbers
Numbers are generally treated as numeric objects (such as double precision real numbers). Meaning, even if you see a '1' it is represented behind the scene as a numeric object like '1.00'. 
If you explicitly want an integer, use the L suffix. Entering '1' gives you a numeric object, while entering '1L' gives an integer object.   
Inf represents infinity. 
NaN represents an undefined value such as 0/0

## Attributes
R Objects can have many attributes (like metadata) to help describe the object. 
* Names, dimnames
* dimensions (matrices, arrays)
* class (integer, numeric)
* length
* other user-defined attributes/metadata

Attributes can be accessed using attributes() function. 

## Creating Vectors
The c() function can be used to create vectors of objects by concatenating things together. 
```{r}
x <- c(0.5, 0.6)       # numeric
x <- c(TRUE, FALSE)    # logical
x <- c(T, F)           # logical
x <- c("a", "b", "c")  # character
x <- 9:29              # integer
x <- c(1+0i, 2+4i)     # complex
x <- vector("numeric", length = 10) # initialize vector
```

## Mixing Objects
Sometimes we mix objects (on accident or on purpose).  
When different objects are mixed in a vector, they are *coerced* to be the same class. Sometimes this works the way you expect others, not. 

## Explicit Coercion
You can explicitly coerce between classes using the as.* function
```{r}
x <- 0:6
class(x)
as.numeric(x)
as.logical(x)
as.character(x)
```
Sometimes, R doesn't know what to do and will return NA. 

## Matrices 
Vectors with a *dimension* attribute. The dimension attribute is an integer vector of length 2
```{r}
m <- matrix (nrow = 2, ncol = 3)
dim(m)
attributes(m)
```
Matrics are constructed *column-wise*, so entries can be thought of starting in the "upper left" corner and running down the columns. 
```{r}
m <- matrix(1:6,nrow = 2, ncol = 3)
m
```
Can also be created directly from vectors by adding a dimension attribute. 
```{r}
m <- 1:10
m
dim(m) <- c(2,5)
m
```
Can be created with *column-binding* or *row-binding*
```{r}
x <- 1:3
y <- 10:13
cbind(x,y)
rbind(x,y)
```

## Lists
Special vectors that contain different elements of different classes.  
Can be explicitly created with the list() function. 
```{r}
x <- list(1, "a", TRUE, 1+4i)
x
```
Can also create an empty list of length with the vector() function.  
```{r}
x <- vector("list", length = 5)
```

## Factors
Used to represent categorical data that is unordered or ordered. Akin to integer vector where each integer has a label. Important for statical modeling and treated special in functions like lm() and glm().  
Using factors with labels is better than using integers because factors are self-describing. 
Factor with values "Male" and "Female" is better than variable with values 1 and 2. 
Create with factor() function
```{r}
x <- factor(c("yes", "yes", "no", "yes", "yes"))
x
table(x)
unclass(x)
```
Often, factors are automatically created when reading data with function like data.table().  
Order of levels can be set using levels argument to factor(). Often important in linear modeling with first level as baseline. 

## Missing Values
